Exploratory data analysis for high-dimensional biological datasets

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
update fluentGenomics chapter

## Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}

```{r setup, include = FALSE}
library(fluentGenomics)
dir <- system.file("extdata", package="macrophage")

library(tximeta)
makeLinkedTxome(
  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
  source="Gencode",
  organism="Homo sapiens",
  release="29",
  genome="GRCh38",
  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
  write=FALSE
)

knitr::opts_chunk$set(
  fig.align = "center"
)
```

@@

# Introduction

@@

```{r workflow, fig.cap = "(ref:workflow)", fig.align='center', out.width="\\textwidth", echo = FALSE}
knitr::include_graphics("workflow.png")
```

@@

## Experimental Data

@@

# Import Data as a *SummarizedExperiment* {#se}

## Using *tximeta* to import RNA-seq quantification data

@@

```{r setdir}
dir <- system.file("extdata", package="macrophage")
```

@@

```{r coldata-rna}
library(dplyr)
library(readr)
colfile <- file.path(dir, "coldata.csv")
coldata <- read_csv(colfile) %>%
  dplyr::select(
    names,
    id = sample_id,
    line = line_id,
    condition = condition_name
  ) %>%
  dplyr::mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    line = factor(line),
    condition = relevel(factor(condition), "naive")
  )
coldata
```

@@

```{r tximeta-run}
suppressPackageStartupMessages(library(SummarizedExperiment))
library(tximeta)
se <- tximeta(coldata, dropInfReps=TRUE)
se
```
<!-- should we describe the data structure in more detail, or use a figure -->

@@

```{r gse}
gse <- summarizeToGene(se)
```

@@

## Importing ATAC-seq data as a *SummarizedExperiment* object {#atac}

@@

# Model assays

## RNA-seq differential gene expression analysis

@@

```{r setup-deseq}
library(DESeq2)
dds <- DESeqDataSet(gse, ~line + condition)
# filter out lowly expressed genes
# at least 10 counts in at least 6 samples
keep <- rowSums(counts(dds) >= 10) >= 6
dds <- dds[keep,]
```

@@

```{r fit-model}
dds <- DESeq(dds)
```

@@

```{r results-DFrame}
res <- results(dds,
               contrast=c("condition","IFNg","naive"),
               lfcThreshold=1, alpha=0.01)
```

@@

```{r ma-plot, fig.cap="(ref:maplot)" }
summary(res)
DESeq2::plotMA(res, ylim=c(-10,10))
```

@@

```{r results-GRanges}
suppressPackageStartupMessages(library(plyranges))
de_genes <- results(dds,
                    contrast=c("condition","IFNg","naive"),
                    lfcThreshold=1,
                    format="GRanges") %>%
  names_to_column("gene_id")
de_genes
```

@@

```{r de-genes}
de_genes <- de_genes %>%
  filter(padj < 0.01) %>%
  select(gene_id, de_log2FC = log2FoldChange, de_padj = padj)
```

@@

```{r not-de-genes}
other_genes <- results(dds,
                       contrast=c("condition","IFNg","naive"),
                       lfcThreshold=1,
                       altHypothesis="lessAbs",
                       format="GRanges") %>%
  filter(padj < 0.01) %>%
  names_to_column("gene_id") %>%
  dplyr::select(gene_id,
                de_log2FC = log2FoldChange,
                de_padj = padj)
```

## ATAC-seq peak differential abundance analysis

@@

<!-- We can check the standard deviation over mean plot, to assess for any -->
<!-- systematic trends: -->

<!-- ```{r, eval = FALSE} --> <!-- library(ggplot2) --> <!-- mv <-
data.frame(rmu = rowMeans(assay(atac)), --> <!-- rvar = rowVars(assay(atac)))
--> <!-- ggplot(data =mv , aes(x = rmu, y = sqrt(rvar))) + --> <!-- geom_hex()
--> <!-- ``` -->

@@

```{r limma, eval = FALSE}
library(limma)
design <- model.matrix(~donor + condition, colData(atac))
fit <- lmFit(assay(atac), design)
fit <- eBayes(fit)
idx <- which(colnames(fit$coefficients) == "conditionIFNg")
tt <- topTable(fit, coef=idx, sort.by="none", n=nrow(atac))
```

@@

```{r peaks-tidy, eval = FALSE}
atac_peaks <- rowRanges(atac) %>%
  remove_names() %>%
  mutate(
    da_log2FC = tt$logFC,
    da_padj = tt$adj.P.Val
  ) %>%
  set_genome_info(genome = "hg38")

@@

```{r load-peaks}
library(fluentGenomics)
peaks
```

# Integrate ranges

## Finding overlaps with *plyranges*

@@

```{r filter-peaks}
da_peaks <- peaks %>%
  filter(da_padj < 0.01)
```

@@

## Down sampling non-differentially expressed genes

@@

```{r slice-example}
size <- length(de_genes)
slice(other_genes, sample.int(n(), size))
```

@@

```{r boot-set-01}
# set a seed for the results
set.seed(2019-08-02)
boot_genes <- replicate(10,
                        slice(other_genes, sample.int(n(), size)),
                        simplify = FALSE)
```

@@

```{r boot-set-02}
boot_genes <- bind_ranges(boot_genes, .id = "resample")
```

@@

```{r combine-results}
all_genes <- bind_ranges(
  de=de_genes,
  not_de = boot_genes,
  .id="origin"
) %>%
  mutate(
    origin = factor(origin, c("not_de", "de")),
    resample = ifelse(is.na(resample), 0L, as.integer(resample))
  )
all_genes
```

## Expanding genomic coordinates around the transcription start site

@@

```{r resize-01}
all_genes <- all_genes %>%
  anchor_5p() %>%
  mutate(width = 1)
```

@@

```{r resize-02}
all_genes <- all_genes %>%
  anchor_center() %>%
  mutate(width=2*1e4)
```

## Use overlap joins to find relative enrichment

@@

```{r olap-join}
genes_olap_peaks <- all_genes %>%
  join_overlap_left(da_peaks)
genes_olap_peaks
```

@@

```{r reduce-ex01}
gene_peak_max_lfc <- genes_olap_peaks %>%
  group_by(gene_id, origin)  %>%
  reduce_ranges_directed(
    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
    peak_max_lfc = max(abs(da_log2FC))
  )
```

@@

```{r boxplot, fig.cap = "(ref:boxplot)"}
library(ggplot2)
gene_peak_max_lfc %>%
  filter(peak_count > 0) %>%
  as.data.frame() %>%
  ggplot(aes(origin, peak_max_lfc)) +
  geom_boxplot()
```

@@

```{r summarize-ex01}
origin_peak_lfc <- genes_olap_peaks %>%
  group_by(origin) %>%
  summarize(
    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
    lfc1_peak_count =sum(abs(da_log2FC) > 1, na.rm=TRUE)/ n_distinct(resample),
    lfc2_peak_count = sum(abs(da_log2FC) > 2, na.rm=TRUE)/ n_distinct(resample)
  )
origin_peak_lfc
```

@@

```{r pivot-enrich}
origin_peak_lfc %>%
  as.data.frame() %>%
  tidyr::pivot_longer(cols = -origin) %>%
  tidyr::pivot_wider(names_from = origin, values_from = value) %>%
  mutate(enrichment = de / not_de)
```

@@

```{r reduce-summarize}
genes_olap_peaks %>%
  group_by(gene_id, origin, resample) %>%
  reduce_ranges_directed(
    lfc1 = sum(abs(da_log2FC) > 1, na.rm=TRUE),
    lfc2 = sum(abs(da_log2FC) > 2, na.rm=TRUE)
  ) %>%
  group_by(origin) %>%
  summarize(
    lfc1_gene_count = sum(lfc1 > 0) / n_distinct(resample),
    lfc1_peak_count = sum(lfc1) / n_distinct(resample),
    lfc2_gene_count = sum(lfc2 > 0) / n_distinct(resample),
    lfc2_peak_count = sum(lfc2) / n_distinct(resample)
  )
```

@@

```{r count-fn}
count_if_above_threshold <- function(var, thresholds) {
  lapply(thresholds, function(.) sum(abs(var) > ., na.rm = TRUE))
}
```

@@

```{r thresholds}
thresholds <- da_peaks %>%
  mutate(abs_lfc = abs(da_log2FC)) %>%
  with(
    seq(min(abs_lfc), max(abs_lfc), length.out = 100)
  )
```

@@

```{r reduce-ex02}
genes_peak_all_thresholds <- genes_olap_peaks %>%
  group_by(gene_id, origin, resample) %>%
  reduce_ranges_directed(
    value = count_if_above_threshold(da_log2FC, thresholds),
    threshold = list(thresholds)
  )
genes_peak_all_thresholds
```

@@

```{r expand-summarize}
origin_peak_all_thresholds <- genes_peak_all_thresholds %>%
  expand_ranges() %>%
  group_by(origin, threshold) %>%
  summarize(
    gene_count = sum(value > 0) / n_distinct(resample),
    peak_count = sum(value) / n_distinct(resample)
  )
origin_peak_all_thresholds
```

@@

```{r line-chart, fig.cap = "(ref:linechart)"}
origin_threshold_counts <- origin_peak_all_thresholds %>%
  as.data.frame() %>%
  tidyr::pivot_longer(cols = -c(origin, threshold),
                      names_to = c("type", "var"),
                      names_sep = "_",
                      values_to = "count") %>%
  select(-var)

origin_threshold_counts %>%
  filter(type == "peak") %>%
  tidyr::pivot_wider(names_from = origin, values_from = count) %>%
  mutate(enrichment =  de / not_de) %>%
  ggplot(aes(x = threshold, y = enrichment)) +
  geom_line() +
  labs(x = "logFC threshold", y = "Relative Enrichment")
```

@@

```{r line-chart2, fig.cap = "(ref:linechart2)"}
origin_threshold_counts %>%
  ggplot(aes(x = threshold,
             y = count + 1,
             color = origin,
             linetype = type)) +
  geom_line() +
  scale_y_log10()
```

# Discussion

@@

# Software Availability

The workflow materials, including this article can be fully reproduced
following the instructions found at the Github repository
[sa-lee/fluentGenomics](https://github.com/sa-lee/fluentGenomics). Moreover,
the development version of the workflow and all downstream dependencies can be
installed using the `BiocManager` package by running:

@@

This article and the analyses were performed with R [@baser] using the
*rmarkdown* [@rmarkdown], and *knitr* [@knitr; @xie2015] packages.

## Session Info

```{r}
sessionInfo()
```

## Author Contributions

All authors contributed to the writing and development of the workflow.

## Competing interests

The authors declare that they have no competing interests.

## Funding

SL is supported by an Australian Government Research Training Program (RTP)
scholarship with a top up scholarship from CSL Limited.

MIL's contribution is supported by NIH grant R01 HG009937.

__I confirm that the funders had no role in study design, data collection and
analysis, decision to publish, or preparation of the manuscript.__

@@

# References

>>>>

# Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}

```{r fluentGenomics-init, echo = FALSE, cache = FALSE, include = FALSE}
knitr::read_chunk(here::here("scripts", "fluentGenomics.R"))
```

```{r setup, include = FALSE}
```

@@

## Introduction

@@

```{r workflow, fig.cap = "(ref:workflow)"}
knitr::include_graphics(here::here("img", "workflow.png"))
```

@@

### Experimental Data

@@

## Import Data as a *SummarizedExperiment* {#se}

### Using *tximeta* to import RNA-seq quantification data

@@

```{r setdir, echo = TRUE}
```

@@

```{r coldata-rna, echo = TRUE}
library(dplyr)
library(readr)
colfile <- file.path(dir, "coldata.csv")
coldata <- read_csv(colfile) %>%
  dplyr::select(
    names,
    id = sample_id,
    line = line_id,
    condition = condition_name
  ) %>%
  dplyr::mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    line = factor(line),
    condition = relevel(factor(condition), "naive")
  )
coldata
```

@@

```{r tximeta-run, echo = TRUE}
```
<!-- should we describe the data structure in more detail, or use a figure -->

@@

```{r gse, echo = TRUE}
```

@@

### Importing ATAC-seq data as a *SummarizedExperiment* object {#atac}

@@

## Model assays

### RNA-seq differential gene expression analysis

@@

```{r setup-deseq, echo = TRUE}
```

@@

```{r fit-model, echo = TRUE}
```

@@

```{r results-DFrame, echo = TRUE}
```

@@

```{r ma-plot, fig.cap="(ref:maplot)", echo = TRUE}
```

@@

```{r results-GRanges, echo = TRUE}
```

@@

```{r de-genes, echo = TRUE}
```

@@

```{r not-de-genes, echo = TRUE}
```

### ATAC-seq peak differential abundance analysis

@@

@@

```{r limma, eval = FALSE, echo = TRUE}
library(limma)
design <- model.matrix(~donor + condition, colData(atac))
fit <- lmFit(assay(atac), design)
fit <- eBayes(fit)
idx <- which(colnames(fit$coefficients) == "conditionIFNg")
tt <- topTable(fit, coef=idx, sort.by="none", n=nrow(atac))
```

@@

```{r peaks-tidy, eval = FALSE, echo = TRUE}
atac_peaks <- rowRanges(atac) %>%
  remove_names() %>%
  mutate(
    da_log2FC = tt$logFC,
    da_padj = tt$adj.P.Val
  ) %>%
  set_genome_info(genome = "hg38")

@@

```{r load-peaks, echo = TRUE}
```

## Integrate ranges

### Finding overlaps with *plyranges*

@@

```{r filter-peaks, echo = TRUE}
```

@@

### Down sampling non-differentially expressed genes

@@

```{r slice-example, echo = TRUE}
```

@@

```{r boot-set-01, echo = TRUE}
```

@@

```{r boot-set-02, echo = TRUE}
```

@@

```{r combine-results, echo = TRUE}
```

### Expanding genomic coordinates around the transcription start site

@@

```{r resize-01, echo = TRUE}
```

@@

```{r resize-02, echo = TRUE}
```

### Use overlap joins to find relative enrichment

@@

```{r olap-join, echo = FALSE}
```

@@

```{r reduce-ex01, echo = FALSE}
```

@@

```{r boxplot, fig.cap = "(ref:boxplot)", echo = TRUE}
```

@@

```{r summarize-ex01, echo = TRUE}
```

@@

```{r pivot-enrich, echo = TRUE}
```

@@

```{r reduce-summarize, echo = TRUE}
```

@@

```{r count-fn, echo = TRUE}
```

@@

```{r thresholds, echo = TRUE}
```

@@

```{r reduce-ex02, echo = TRUE}
```

@@

```{r expand-summarize, echo = TRUE}
```

@@

```{r line-chart, fig.cap = "(ref:linechart)", echo = TRUE}
```

@@

```{r line-chart2, fig.cap = "(ref:linechart2)", echo = TRUE}
```

## Discussion

@@

## Software Availability

The workflow materials can be fully reproduced
following the instructions found at the Github repository
[sa-lee/fluentGenomics](https://github.com/sa-lee/fluentGenomics). Moreover,
the development version of the workflow and all downstream dependencies can be
installed using the `BiocManager` package by running:

@@

@@
